---
output:
  pdf_document: default
  html_document: default
---
# Value of Plays Using Run Expectancy


**Learning objectives:**

- Run Expectancy
- Batting Lineup
- Run values by bases
- Value of stealing

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library("abdwr3edata")
library("tidyverse")

crcblue <- "#2905A1"
crc_fc <- c("#2905a1", "#e41a1c", "#4daf4a", "#984ea3")
```

## Run Expectancy Matrix

Run Expectancy Matrix- Average number of runs scored in the remainder of the inning for each combination of runners and outs

-   8 possible arrangements of runners
-   24 possible arrangements of runners and outs

Use matrix to find change in expected run value

**What is the relative benefit of different batting plays?**


## Runs Scored in the Remainder of the Inning

Potential for runs- avg run scored in remainder of the inning for each combination of runners and outs

-   Represents opportunity to create runs during an inning
-   Baseline to measure player contributions against

Runs scored in the Remainder Of the Inning (roi):
$$runs_{\text{roi}} = runs_{\text{Total in inning}} - runs_{\text{So far in inning}}$$

```{r, echo = FALSE, eval = TRUE}
retro2016 <- retro2016 |> 
  mutate(
    runs_before = away_score_ct + home_score_ct,
    half_inning = paste(game_id, inn_ct, bat_home_id),
    runs_scored = 
      (bat_dest_id > 3) + (run1_dest_id > 3) + 
      (run2_dest_id > 3) + (run3_dest_id > 3)
  )

half_innings <- retro2016 |>
  group_by(half_inning) |>
  summarize(
    outs_inning = sum(event_outs_ct), 
    runs_inning = sum(runs_scored),
    runs_start = first(runs_before),
    max_runs = runs_inning + runs_start
  )

retro2016 <- retro2016 |>
  inner_join(half_innings, by = "half_inning") |>
  mutate(runs_roi = max_runs - runs_before)
```


## Creating the Matrix

Consider plays with change in:

-   runners on bases

-   number of outs

-   runs scored

Get states (baserunners and outs) to identify changes and filter

Only consider half-innings with 3 outs [small bias]

```{r, echo = FALSE, eval = TRUE}
retro2016 <- retro2016 |>
  mutate(
    bases = paste0(
      if_else(base1_run_id == "", 0, 1),
      if_else(base2_run_id == "", 0, 1),
      if_else(base3_run_id == "", 0, 1)
    ),
    state = paste(bases, outs_ct)
  )

retro2016 <- retro2016 |>
  mutate(
    is_runner1 = as.numeric(
      run1_dest_id == 1 | bat_dest_id == 1
    ),
    is_runner2 = as.numeric(
      run1_dest_id == 2 | run2_dest_id == 2 | 
        bat_dest_id == 2
    ),
    is_runner3 = as.numeric(
      run1_dest_id == 3 | run2_dest_id == 3 |
        run3_dest_id == 3 | bat_dest_id == 3
    ),
    new_outs = outs_ct + event_outs_ct,
    new_bases = paste0(is_runner1, is_runner2, is_runner3),
    new_state = paste(new_bases, new_outs)
  )

changes2016 <- retro2016 |> 
  filter(state != new_state | runs_scored > 0)

changes2016_complete <- changes2016 |>
  filter(outs_inning == 3)
```

Run expectancy (exp. number of runs scored in the remainder of the inning):

```{r, include = FALSE}
erm_2016 <- changes2016_complete |> 
  group_by(bases, outs_ct) |>
  summarize(mean_run_value = mean(runs_roi))
```


```{r, echo = FALSE}
ggplot(erm_2016, aes(x = factor(outs_ct), y = bases, fill = mean_run_value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(mean_run_value, 2)), size = 4, color = "black") +
  scale_fill_gradientn(colors = heat.colors(10), name = "Run Value")
```


```{r, echo = FALSE}
base_order <- c("000", "100", "010", "001", "110", "101", "011", "111")

erm_2016 <- erm_2016 |>
  mutate(bases = factor(bases, levels = base_order))

ggplot(erm_2016, aes(x = bases, y = mean_run_value, color = outs_ct)) +
  geom_point(size = 3) +
  labs(x = "Bases", y = "Mean Run Value", color = "Outs") 
  
```


Change over time:


```{r, echo = FALSE}
base_levels <- c("000", "100", "010", "001", "110", "101", "011", "111")

erm_2002 <- tibble(
  bases = base_levels,
  "OLD=0" = c(.51, 1.40, 1.14,  1.96, .90, 1.84, 1.51, 2.33), 
  "OLD=1" = c(.27,  .94,  .68,  1.36, .54, 1.18,  .94, 1.51),
  "OLD=2" = c(.10,  .36,  .32,   .63, .23,  .52,  .45,  .78)
) |>
  mutate(bases = base_levels)

erm_2016_reorg <- changes2016_complete |>
  group_by(bases, outs_ct) |>
  summarize(mean_run_value = mean(runs_roi), .groups = "drop") |>
  mutate(bases = factor(bases, levels = base_levels)) |>
  pivot_wider(
    names_from = outs_ct, 
    values_from = mean_run_value, 
    names_prefix = "NEW="
  )

combined <- bind_cols(erm_2016_reorg, erm_2002 |> select(-bases)) |>
  pivot_longer(
    cols = starts_with("NEW=") | starts_with("OLD="),
    names_to = "source_outs",
    values_to = "mean_run_value"
  ) |>
  separate(source_outs, into = c("source", "outs"), sep = "=") |>
  mutate(outs = factor(outs, levels = c("0", "1", "2")))

ggplot(combined, aes(x = outs, y = bases, fill = mean_run_value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(mean_run_value, 2)), size = 4) +
  scale_fill_gradientn(colors = heat.colors(10), name = "Run Value") +
  facet_wrap(~ source, ncol = 2, labeller = labeller(source = c("NEW" = "New (2016)", "OLD" = "Old (2002)"))) 
  
```

Few changes in average run scoring tendencies of MLB teams 2002-2016


## Measuring Success of a Batting Play

Run Value- value of plate appearance; state will change and run expectancy will be updated

$$\text{Run Value} = Runs_{\text{New state}} - Runs_{\text{Old state}} + Runs_{\text{Scored on Play}}$$


Data handling:

-   Match expected run values for beginning of each plate appearance by matching bases and outs to run expectancy matrix

-   Match new bases and new outs to run expectancy matrix

-   Set run values with 3 outs to 0

-   End with: run expectancy of current state, run expectancy of new state, difference of run expectancy of current state and new state plus run scored

```{r, echo = FALSE, eval = TRUE}
retro2016 <- retro2016 |>
  left_join(erm_2016, join_by("bases", "outs_ct")) |>
  rename(rv_start = mean_run_value) |>
  left_join(
    erm_2016, 
    join_by(new_bases == bases, new_outs == outs_ct)
  ) |>
  rename(rv_end = mean_run_value) |>
  replace_na(list(rv_end = 0)) |>
  mutate(run_value = rv_end - rv_start + runs_scored)
```


## José Altuve

Isolate to batting plays where Altuve was the hitter

```{r, echo = FALSE, eval = TRUE}
library(Lahman)

altuve_id <- People |> 
  filter(nameFirst == "Jose", nameLast == "Altuve") |>
  pull(retroID)

altuve <- retro2016 |> 
  filter(
    bat_id == altuve_id,
    bat_event_fl == TRUE
  )
```


**How did Altuve do in his first three plate appearances in the 2016 season with the Houston Astros?**

```{r}
altuve |> 
  select(state, new_state, run_value) |>
  slice_head(n = 3)
```

        
**2 primary questions:**

-    What were the runner/outs situations for the player's plate appearances?

-    What did the batter do with these opportunities to score runs? 



**Runner states to understand opportunities**

```{r}
altuve |> 
  group_by(bases) |> 
  summarize(N = n())
```

Most often batting with bases empty or runner only on first

```{r, echo = FALSE}
ggplot(altuve, aes(bases, run_value, color = run_value)) +
  geom_jitter(width = 0.25, alpha = 0.5) +
  geom_hline(yintercept = 0, color = "red") +
  xlab("Runners on base") +
  scale_color_gradient(low = "#002D62", high = "#EB6E1F") +
  annotate("rect",
           xmin = as.numeric(factor("000", levels = unique(altuve$bases))) - 0.3,
           xmax = as.numeric(factor("000", levels = unique(altuve$bases))) + 0.3,
           ymin = 0.95, ymax = 1.05,
           color = "#F4911E", fill = NA, linewidth = 1)
```


Performance with these opportunities:

-   When bases empty- range of runs was small

-   Other situations- much more variation



**Total run production for 2016**

Number of opportunities and sum of run values for each situation

```{r}
runs_altuve <- altuve |> 
  group_by(bases) |>
  summarize(
    PA = n(),
    total_run_values = sum(run_value)
  )
runs_altuve
```

-   Total contribution at bat with bases empty: 10.10 runs

-   Didn't do well with runners in scoring position

Total contribution for 2016 season (run expectancy change over 24 base/out states):

```{r}
runs_altuve |> 
  summarize(RE24 = sum(total_run_values))
```


## Opportunity and Success for All Hitters

Comparing run estimates to evaluate player batting effectiveness

```{r, echo = FALSE}
retro2016_bat <- retro2016 |> 
  filter(bat_event_fl == TRUE)

run_exp <- retro2016_bat |> 
  group_by(bat_id) |>
  summarize(
    RE24 = sum(run_value),
    PA = length(run_value),
    runs_start = sum(rv_start)
  )
```

**Different players have different opportunities to create runs (batting order)**

Performance by season:

-   Total number of plate appearances

-   Sum of runs potentials

-   Sum of the run values



Focus on batters 400+ plate appearances

```{r}
run_exp_400 <- run_exp |> 
  filter(PA >= 400)
run_exp_400 |>
  slice_head(n = 6)
```

**Is there a relationship between batters' opportunities and their success in converting opportunities to runs?**

```{r, echo = FALSE}
plot1 <- ggplot(run_exp_400, aes(runs_start, RE24)) +
  geom_point() + 
  geom_smooth() +
  geom_hline(yintercept = 0, color = "red")
plot1
```

Batters with larger runs start values tend to have larger run contributions but there is a wide spread

```{r, echo = FALSE}
run_exp_400 <- run_exp_400 |>
  inner_join(People, by = c("bat_id" = "retroID"))
library(ggrepel)
plot1 + 
  geom_text_repel(
    data = filter(run_exp_400, RE24 >= 40), 
    aes(label = nameLast)
  ) +
  geom_text_repel(
    data = filter(run_exp_400, nameLast == "Altuve"), 
    aes(label = nameLast),
    nudge_y = -14,                # Offset vertically
    nudge_x = 16,
    segment.color = "black",   # Line color to point
    segment.size = 0.5,        # Line thickness
    fontface = "bold",
    color = "#F4911E"
  )
```


## Position in the Batting Lineup

Evaluating run value criterion in relation to batting position

Note: Using frequency to define batting position

```{r, echo = FALSE}
regulars <- retro2016 |>
  inner_join(run_exp_400, by = "bat_id")

positions <- regulars |> 
  group_by(bat_id, bat_lineup_id) |>
  summarize(N = n()) |> 
  arrange(desc(N)) |> 
  mutate(position = first(bat_lineup_id))

run_exp_400 <- run_exp_400 |>
  inner_join(positions, by = "bat_id")
```

```{r, echo = FALSE}
run_exp_400$position <- as.factor(run_exp_400$position)
ggplot(run_exp_400, aes(runs_start, RE24, label = position, color = position)) +
  geom_text() +
  geom_hline(yintercept = 0, color = "red") +
  geom_point(
    data = filter(run_exp_400, bat_id == altuve_id),
    size = 4, shape = 16, color = crcblue
  ) +
  scale_color_brewer(palette = "Set1")
  
```

First and second batters create fewer runs than third and fourth

Best hitters generally bat third, fourth, and fifth:

```{r, echo = FALSE}
run_exp_400_posnear3 <- run_exp_400 %>%
  filter(position == c(3, 4, 5))

ggplot(run_exp_400_posnear3, aes(runs_start, RE24, label = position, color = position)) +
  geom_text() +
  geom_hline(yintercept = 0, color = "red") +
  geom_point(
    data = filter(run_exp_400, bat_id == altuve_id),
    size = 4, shape = 16, color = crcblue
  ) +
  scale_color_brewer(palette = "Set1")
```

Though some fourth (and fifth) position hitters have mediocre run values

Altuve is in the top hitters in terms of creating runs


## Value of a home run

Batting average gives equal value to the four possible base hits.

```{r, echo = FALSE}
home_runs <- retro2016 |> 
  filter(event_cd == 23)
```
```{r}
home_runs |>
  select(state) |>
  table()
```

```{r, echo = FALSE}
hr_statevals <- home_runs |>
  count(state) |>
  mutate(
    bases = substr(state, 1, 3),
    outs = substr(state, 5, 5)
  )

# Step 2: Create heatmap
ggplot(hr_statevals, aes(x = outs, y = bases, fill = n)) +
  geom_tile(color = "white") +
  geom_text(aes(label = n), size = 4) +
  scale_fill_gradientn(colors = heat.colors(10), name = "Count")
```


```{r}
hr_statevals <- home_runs |>
  select(state) |>
  table() |>
  prop.table() |>
  round(3)
hr_statevals
```

```{r, echo = FALSE}
hr_statevals_df <- home_runs |>
  count(state) |>
  mutate(prop = round(n / sum(n), 3)) |>
  mutate(
    bases = substr(state, 1, 3),
    outs = substr(state, 5, 5)
  )

ggplot(hr_statevals_df, aes(x = outs, y = bases, fill = prop)) +
  geom_tile(color = "white") +
  geom_text(aes(label = prop), size = 4) +
  scale_fill_gradientn(colors = heat.colors(10), name = "Proportion")
```

Assess by bases:

```{r, echo = FALSE}
state_prefixes <- sub("^(\\d{3}) \\d$", "\\1", names(hr_statevals))

tapply(hr_statevals, state_prefixes, sum)
```

Over half of the home runs are hit with no runners on base

**What is the run value of a home run?**

```{r}
mean_hr <- home_runs |>
  summarize(mean_run_value = mean(run_value))
mean_hr
```

```{r, echo = FALSE}
ggplot(home_runs, aes(run_value)) +
  geom_histogram() + 
  geom_vline(
    data = mean_hr, aes(xintercept = mean_run_value), 
    color = "red", linewidth = 1.5
  ) +
  annotate(
    "text", 1.7, 2000, 
    label = "Mean Run\nValue", color = "red"
  )
```

Most home runs occur with bases empty (1)

**Which runner/out situations lead to the most valuable home runs?**

```{r}
home_runs |> 
  arrange(desc(run_value)) |>
  select(state, new_state, run_value) |>
  slice_head(n = 1)
```

Bases loaded, 2 outs


## Value of a single

Final state of a home run: bases empty

Final sate of a single: depends on movement of runners on bases

```{r, echo = FALSE}
singles <- retro2016 |> 
  filter(event_cd == 20)

mean_singles <- singles |>
  summarize(mean_run_value = mean(run_value))

ggplot(singles, aes(run_value)) + 
  geom_histogram(bins = 40) +
  geom_vline(
    data = mean_singles, color = "red", 
    aes(xintercept = mean_run_value), linewidth = 1.5
  ) +
  annotate(
    "text", 0.8, 4000, 
    label = "Mean Run\nValue", color = "red"
  )
```


```{r}
hr_single <- singles |>
  select(state) |>
  table()
hr_single
```

```{r, echo = FALSE}
hr_df <- as.data.frame(hr_single)
colnames(hr_df) <- c("state", "count")

hr_df <- hr_df |>
  separate(state, into = c("base", "out"), sep = " ")

hr_matrix <- hr_df |>
  pivot_wider(names_from = out, values_from = count, values_fill = 0)

matrix_data <- as.matrix(hr_matrix[,-1])
rownames(matrix_data) <- hr_matrix$base

melted_matrix <- hr_matrix |>
  pivot_longer(
    cols = -base,
    names_to = "out",
    values_to = "count"
  )

ggplot(melted_matrix, aes(x = out, y = base, fill = count)) +
  geom_tile(color = "white") +
  geom_text(aes(label = count), color = "black") +
  scale_fill_gradientn(colors = heat.colors(10)) +
  theme_minimal()
```



By bases: 

```{r, echo = FALSE}
state_prefixes <- sub("^(\\d{3}) \\d$", "\\1", names(hr_single))

tapply(hr_single, state_prefixes, sum)
```

Most singles occur with bases empty


**What is the most valuable single from the run value perspective?**

```{r}
singles |> 
  arrange(desc(run_value)) |>
  select(state, new_state, run_value) |>
  slice_head(n = 1)
```

The Baseball-Reference notes the center fielder made an error on fielding the single that allowed all three runners to score and the batter to reach third

Smallest run values:

```{r}
singles |> 
  arrange(run_value) |>
  select(state, new_state, run_value) |>
  slice(1)
```

A runner on second was hit by the ball in play and called out, creating a negative run value

**The benefit of the single depends on the advancement of the runners**


## Value of Base Stealing

**Is there a net benefit to attempting to steal a base?**

```{r, echo = FALSE}
stealing <- retro2016 |> 
  filter(event_cd %in% c(4, 6))

stealing |> 
  mutate(event_cd = recode(event_cd, '4' = 'Stolen Base', '6' = 'Caught Stealing')) |>
  group_by(event_cd) |> 
  summarize(N = n()) |> 
  mutate(pct = N / sum(N))
```

**What are common runners/outs situations for attempting a stolen base?**

```{r}
stealing |> 
  group_by(state) |> 
  summarize(N = n())
```

Most common: runner only on first

The corresponding run value reflects the success of the attempt.

```{r, echo = FALSE}
ggplot(stealing, aes(run_value, fill = factor(event_cd))) + 
  geom_histogram() +
  scale_fill_manual(
    name = "event_cd",
    values = crc_fc,
    labels = c("Stolen Base (SB)", "Caught Stealing (CS)")
  )
```


**Focus on stolen base attempts**

State of 100 1
```{r, echo = FALSE}
stealing_1001 <- stealing |> 
  filter(state == "100 1")
```

```{r, echo = FALSE}
stealing_1001 |> 
  mutate(event_cd = recode(event_cd, '4' = 'Stolen Base', '6' = 'Caught Stealing')) |> 
  group_by(event_cd) |> 
  summarize(N = n()) |> 
  mutate(pct = N / sum(N))
```

Runner stealing success = 70.3%


Broken down by New States:

```{r, echo = FALSE}
stealing_1001 |> 
  group_by(new_state) |> 
  summarize(N = n()) |> 
  mutate(pct = N / sum(N))
```

-   457 occurrences of successfully advancing to second

-   39 occurrences of successfully advancing to third


**Value of attempting stolen bases in this situation**

```{r}
stealing_1001 |> 
  summarize(Mean = mean(run_value))
```

Stolen bases are worthwhile but overall value is 0.007 runs/attempt

**BUT** benefit really depends on success/failure and runners/outs situations
