# Career Trajectories

**Learning objectives:**

- Refresh on regression models (`lm`) in R
- Study player career trajectories

## Mickey Mantle's trajectory - Warm up


```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(Lahman)

mantle_id <- People |> 
  filter(nameFirst == "Mickey", nameLast == "Mantle") |>
  pull(playerID)

# deal with missing stats in early dat
batting <- Batting |>
  replace_na(list(SF = 0, HBP = 0))
```

- Define a function to compute OBP = OBS + SLG and Age

```{r}
get_stats <- function(player_id) {
  batting |> 
    filter(playerID == player_id) |>
    inner_join(People, by = "playerID") |>
    mutate(
      birthyear = if_else(
        birthMonth >= 7, birthYear + 1, birthYear
      ),
      Age = yearID - birthyear,
      SLG = (H - X2B - X3B - HR + 2 * X2B + 3 * X3B + 4 * HR) / AB,
      OBP = (H + BB + HBP) / (AB + BB + HBP + SF),
      OPS = SLG + OBP
    ) |>
    select(Age, SLG, OBP, OPS)
}
```

- Finaly we can use this to look at OPS for Micky Mantle:

```{r}
Mantle <- get_stats(mantle_id)
ggplot(Mantle, aes(Age, OPS)) + geom_point() 
```

## Fit to parabola {-}

$$
OPS(\text{Age}) = \text{MAX} - \gamma (\text{Age} - \text{PEAK_AGE})^2
$$
- Instead of fitting this nonlinear model, we transform it into this linear model:

$$
\begin{aligned}
OPS(\text{Age}) &=  A + B (\text{Age}-30) + C (\text{Age} - 30))^2\\
\text{PEAK_AGE}  &= 30 - \frac{B}{2C}\\
\text{MAX} &= A - \frac{B^2}{4C}
\end{aligned}
$$
- This can be fit with `lm` using the formula:

```
OPS ~ I(Age - 30) + I((Age - 30)^2)
```

## Fit (Mantle) {-}

```{r}
fit_model <- function(d) {
  fit <- lm(OPS ~ I(Age - 30) + I((Age - 30)^2), data = d)
  b <- coef(fit)
  Age_max <- 30 - b[2] / b[3] / 2
  Max <- b[1] - b[2] ^ 2 / b[3] / 4
  list(fit = fit, Age_max = Age_max, Max = Max)
}
F2 <- fit_model(Mantle)
F2 |>
  pluck("fit") |>
  coef()
```

- What we really want though:

```{r}
c(F2$Age_max, F2$Max)
```

## Plot it {-}

```{r}
ggplot(Mantle, aes(Age, OPS)) + geom_point() +
  geom_smooth(
    method = "lm", se = FALSE, linewidth = 1.5,
    formula = y ~ poly(x, 2, raw = TRUE)
  ) +
  geom_vline(
    xintercept = F2$Age_max, 
    linetype = "dashed", color = "red"
  ) +
  geom_hline(
    yintercept = F2$Max,
    linetype = "dashed", color = "red"
  ) +
  annotate(
    geom = "text", x = c(29, 20), y = c(0.72, 1.1),
    label = c("Peak age", "Max"), size = 5,
    color = "red"
  )
```

## Full fit summary {-}

```{r}
F2 |> pluck("fit") |> summary()
```

- Note for example $R^2$ about 0.60, meaning 60% of the variation in OPS is 'explained' by the curve. 


## Comparing Trajectories

- Goal for this section is to compare trajectories of 'similar' players

- Uses a measure of similarity developed by Bill James, using both:
    - Position
    - Various statistics (G, AB, H, SO, etc etc )
    

## Setting up the data {-}

- Only players with more then 2000 AB:

```{r}
batting_2000 <- batting |> 
  group_by(playerID) |>
  summarize(AB_career = sum(AB, na.rm = TRUE)) |>
  inner_join(batting, by = "playerID") |>
  filter(AB_career >= 2000)
```

- Add the most played position:

```{r}
Positions <- Fielding |> 
  group_by(playerID, POS) |>
  summarize(Games = sum(G)) |> 
  arrange(playerID, desc(Games)) |> 
  filter(POS == first(POS))

batting_2000 <- batting_2000 |>
  inner_join(Positions, by = "playerID")
```

- Add career statistics

```{r}
my_vars <- c("G", "AB", "R", "H", "X2B", "X3B",
             "HR", "RBI", "BB", "SO", "SB")

C_totals <- batting_2000 |>
  group_by(playerID) |>
  summarize(across(all_of(my_vars), ~ sum(.x, na.rm = TRUE))) |>
  mutate(
    AVG = H / AB,
    SLG = (H - X2B - X3B - HR + 2 * X2B + 3 * X3B + 4 * HR) / AB
  ) |>
  inner_join(Positions, by = "playerID") |>
  mutate(
    Value_POS = case_when(
      POS == "C" ~ 240,
      POS == "SS" ~ 168,
      POS == "2B" ~ 132,
      POS == "3B" ~ 84,
      POS == "OF" ~ 48,
      POS == "1B" ~ 12, 
      TRUE ~ 0
    )
  )
```

- Value_POS are numbers introduced for use in the similarity score by Bill James for positions.

## Compute Similarity Score {-}

- Function finds `number` similar players to a given player based on Bill Jame's similarity

```{r}
similar <- function(p, number = 10) {
  P <- C_totals |> 
    filter(playerID == p)
  C_totals |> 
    mutate(
      sim_score = 1000 -
        floor(abs(G - P$G) / 20) -
        floor(abs(AB - P$AB) / 75) -
        floor(abs(R - P$R) / 10) -
        floor(abs(H - P$H) / 15) -
        floor(abs(X2B - P$X2B) / 5) -
        floor(abs(X3B - P$X3B) / 4) -
        floor(abs(HR - P$HR) / 2) -
        floor(abs(RBI - P$RBI) / 10) -
        floor(abs(BB - P$BB) / 25) -
        floor(abs(SO - P$SO) / 150) -
        floor(abs(SB - P$SB) / 20) - 
        floor(abs(AVG - P$AVG) / 0.001) - 
        floor(abs(SLG - P$SLG) / 0.002) -
        abs(Value_POS - P$Value_POS)
    ) |>
    arrange(desc(sim_score)) |> 
    slice_head(n = number)
}
```

## Example {-}

```{r}
similar(mantle_id, 6)
```

## Compute Age and OPS for all players / seasons {-}



```{r}
batting_2000 <- batting_2000 |> 
  group_by(playerID, yearID) |>  # combine stints
  summarize(
    G = sum(G), AB = sum(AB), R = sum(R),
    H = sum(H), X2B = sum(X2B), X3B = sum(X3B),
    HR = sum(HR), RBI = sum(RBI), SB = sum(SB),
    CS = sum(CS), BB = sum(BB), SH = sum(SH),
    SF = sum(SF), HBP = sum(HBP),
    AB_career = first(AB_career),
    POS = first(POS)
  ) |>
  mutate(
    SLG = (H - X2B - X3B - HR + 2 * X2B + 3 * X3B + 4 * HR) / AB,
    OBP = (H + BB + HBP) / (AB + BB + HBP + SF),
    OPS = SLG + OBP
  ) |>      # add age
  inner_join(People, by = "playerID") |>
  mutate(
    Birthyear = if_else(
      birthMonth >= 7, birthYear + 1, birthYear
    ),
    Age = yearID - Birthyear
  ) |> 
  drop_na(Age)  # Drop NA's 

```

## Fit and plot trajectories {-}

- Note that the fitting is done by ggplot2

```{r}
plot_trajectories <- function(player, n_similar = 5, ncol) { 
  flnames <- unlist(str_split(player, " "))
  
  player <- People |> 
    filter(nameFirst == flnames[1], nameLast == flnames[2]) |>
    select(playerID)

  player_list <- player |>
    pull(playerID) |>
    similar(n_similar) |>
    pull(playerID)
  
  Batting_new <- batting_2000 |> 
    filter(playerID %in% player_list) |>
    mutate(Name = paste(nameFirst, nameLast))
  
    ggplot(Batting_new, aes(Age, OPS)) + 
      geom_smooth(
        method = "lm",
        formula = y ~ x + I(x^2),
        linewidth = 1.5
      ) +
      facet_wrap(vars(Name), ncol = ncol) + 
      theme_bw()
}
```

## Mickey Mantle {-}

```{r}
plot_trajectories("Mickey Mantle", 6, 2)
```

- Take aways?

## Derek Jeter {-}

```{r}
dj_plot <- plot_trajectories("Derek Jeter", 9, 3)
dj_plot
```

- Many differences in curvature and peak age.
- Other take-aways?

## Sumarize by peak Age and curvature {-}

```{r}
library(broom)
data_grouped <- dj_plot$data |> # pull data out of plot 
  group_by(Name)
player_names <- data_grouped |>
  group_keys() |>
  pull(Name)
regressions <- data_grouped |>
  group_split() |>
  map(~lm(OPS ~ I(Age - 30) + I((Age - 30) ^ 2), data = .)) |>
  map(tidy) |>
  set_names(player_names) |>
  bind_rows(.id = "Name")
 

S <- regressions |> 
  group_by(Name) |> 
  summarize(
    b1 = estimate[1],
    b2 = estimate[2],
    Curvature = estimate[3],
    Age_max = round(30 - b2 / Curvature / 2, 1),
    Max = round(b1 - b2 ^ 2 / Curvature / 4, 3)
  )

library(ggrepel)
ggplot(S, aes(Age_max, Curvature, label = Name)) + 
  geom_point() + geom_label_repel()

```


- Q: Is there a general trend of early and sharp peaks?

## General Patterns of Peak Age

- How have peak ages varied across time?

- How does peak age vary with number of at bats?

## Data preperation {-}

- Focus on players that had complete careers 
- Use middle of career to position players careers in time
- Fit them all and extract coefficients

```{r}
not_current_playerID <- People |>
  filter(finalGame < "2021-11-01") |> 
  pull(playerID)
batting_2000 <- batting_2000 |>
  filter(playerID %in% not_current_playerID) 

midcareers <- batting_2000 |>
  group_by(playerID) |>
  summarize(
    Midyear = (min(yearID) + max(yearID)) / 2,
    AB_total = first(AB_career)
  )
batting_2000 <- batting_2000 |>
  inner_join(midcareers, by = "playerID")

batting_2000_grouped <- batting_2000 |> 
  group_by(playerID)

ids <- batting_2000_grouped |>
  group_keys() |>
  pull(playerID)
models <- batting_2000_grouped |>
  group_split() |>
  map(~lm(OPS ~ I(Age - 30) + I((Age - 30)^2), data = .)) |>
  map(tidy) |>
  set_names(ids) |>
  bind_rows(.id = "playerID")

beta_coefs <- models |> 
  group_by(playerID) |> 
  summarize(
    A = estimate[1],
    B = estimate[2],
    C = estimate[3]
  ) |>
  mutate(Peak_age = 30 - B / 2 / C) |>
  inner_join(midcareers, by = "playerID") |>
  inner_join(Positions) |> # for use later!
  rename(Position = POS)
```

## Patterns of peak age over time {-}

```{r}
age_plot <- ggplot(beta_coefs, aes(Midyear, Peak_age)) +
  geom_point(alpha = 0.5) +
  geom_smooth(color = "red", method = "loess") +
  ylim(20, 40) +
  xlab("Mid Career") + ylab("Peak Age")
age_plot
```

- Shows a gradual increase over time (except maybe in last decade?)

## Peak age and career at-bats {-}

```{r}
age_plot +
  aes(x = log2(AB_total)) +
  xlab("Log2 of Career AB")
```

- Longer careers tend to peak at higher Age

- If they peak earlier they also retire earlier?


## Fielding Position

- We already have the fits, just filter by midyear

```{r}
beta_coefs_a <- beta_coefs |>
  filter(Midyear >= 1985, Midyear <= 1995)
```

```{r}
beta_fielders <- beta_coefs_a |>
  filter(
    Position %in% c("1B", "2B", "3B", "SS", "C", "OF")
  ) |> 
  inner_join(People)


ggplot(beta_fielders, aes(Position, Peak_age)) + 
  geom_jitter(width = 0.2) + ylim(20, 40) +
  geom_label_repel(
    data = filter(beta_fielders, Peak_age > 37),
    aes(Position, Peak_age, label = nameLast)
  )
```

- I don't see a clear pattern... thoughts?


## Discussion points

- Is quadradic too naive?  Gaussian process?
- Did anyone do exercises they want to share?



