# Simulation

**Learning objectives:**

- Simulate a half-inning
- Simulate many half-innings
- Simulate a baseball season
- Simulate many baseball seasons

```{r setup_ch_9, message = FALSE, warning = FALSE}
library("abdwr3edata")
library("ggtext")
library("gt")
library("Lahman")
library("tidyverse")

#https://beanumber.github.io/sds192/lab-import.html#Data_in_other_formats
load("data/retro2016.rda") #has retro2016 data frame
```

<details>
<summary>Session Info</summary>

```{r}
sessionInfo()
```

</details>


## Setup

### Retrieve situation states

```{r}
retro2016 <- retro2016 |>
  abdwr3edata::retrosheet_add_states() #from chapter 5
colnames(retro2016[,99:107])
```

### Sum runs and ID half innings

```{r}
half_innings <- retro2016 |>
  mutate(
    runs = away_score_ct + home_score_ct,
    half_inning_id = paste(game_id, inn_ct, bat_home_id)
  ) |>
  group_by(half_inning_id) |>
  summarize(
    outs_inning = sum(event_outs_ct), 
    runs_start = first(runs),
    runs_inning = sum(runs_scored),
    max_runs = runs_start + runs_inning
  )
nrow(half_innings)
```

### Meaningful plays

* retain observations with change in state or runs scored
* complete innings (i.e. three outs)
* ignore: steals, caught stealing, wild pitches, passed balls

```{r}
retro2016_complete <- retro2016 |> 
  mutate(
    half_inning_id = paste(game_id, inn_ct, bat_home_id)
  ) |>
  inner_join(half_innings, join_by(half_inning_id)) |>
  filter(state != new_state | runs_scored > 0) |> 
  filter(outs_inning == 3, bat_event_fl)
print(paste0(round(100*(nrow(retro2016_complete) / nrow(retro2016))), " percent"))
```


### End of innings

> In our definition of the new_state variable, we recorded the runner locations when there were three outs. The runner locations don’t matter, so we recode new_state to always have the value 3 when the number of outs is equal to 3.

```{r}
retro2016_complete <- retro2016_complete |>
  mutate(new_state = str_replace(new_state, "[0-1]{3} 3", "3"))
table(retro2016_complete$new_state)
```

## Transition Matrices

### Transition states

* 24 `state` labels
* 25 `new_state` labels

```{r}
# transition matrix
T_matrix <- retro2016_complete |> select(state, new_state) |> table()
```

```{r, echo = FALSE, eval = TRUE}
T_df <- T_matrix |> 
  as.data.frame() |>
  pivot_wider(names_from = new_state, values_from = Freq)

T_df |>
  gt() |>
  cols_align(align = "center") |>
  data_color(columns = -state,
             palette = "inferno") |>
  tab_header(title = "Play State Transition Matrix",
             subtitle = "2016 Season")
```

<details>
<summary>table code</summary>
```{r, echo = TRUE, eval = FALSE}
T_df <- T_matrix |> 
  as.data.frame() |>
  pivot_wider(names_from = new_state, values_from = Freq)

T_df |>
  gt() |>
  cols_align(align = "center") |>
  data_color(columns = -state,
             palette = "inferno") |>
  tab_header(title = "Play State Transition Matrix",
             subtitle = "2016 Season")
```
</details>

### Absorbing states

> When the inning reaches 3 outs, then it stays at 3 outs, so the probability of staying in this state is 1

```{r}
# probability matrix
P_matrix <- prop.table(T_matrix, 1)
P_matrix <- P_matrix |> rbind("3" = c(rep(0, 24), 1))
```

```{r, echo = FALSE, eval = TRUE}
P_df <- P_matrix |> as.data.frame()
P_df <- cbind(c(as.character(T_df$state), "3"), P_df)
colnames(P_df)[1] <- "state"

P_df |>
  gt() |>
  cols_align(align = "center") |>
  data_color(columns = -state,
             palette = "inferno") |>
  fmt_number(columns = -state,
             decimals = 3) |>
  tab_header(title = "Play State Transition Probability Matrix",
             subtitle = "2016 Season")
```

<details>
<summary>table code</summary>
```{r, echo = TRUE, eval = FALSE}
P_df <- P_matrix |> as.data.frame()
P_df <- cbind(c(as.character(T_df$state), "3"), P_df)
colnames(P_df)[1] <- "state"

P_df |>
  gt() |>
  cols_align(align = "center") |>
  data_color(columns = -state,
             palette = "inferno") |>
  fmt_number(columns = -state,
             decimals = 3) |>
  tab_header(title = "Play State Transition Probability Matrix",
             subtitle = "2016 Season")
```
</details>

### Examples

> we display the transition probabilities starting at the “000 0” state, no runners and no outs below

```{r, echo = FALSE, eval = TRUE}
P_matrix |>
  as_tibble(rownames = "state") |>
  filter(state == "000 0") |>
  pivot_longer(
    cols = -state, 
    names_to = "new_state", 
    values_to = "Prob" 
  ) |>
  filter(Prob > 0) |>
  arrange(desc(Prob)) |>
  gt() |>
  cols_align(align = "center") |>
  data_color(columns = Prob,
             palette = "inferno") |>
  fmt_number(columns = Prob,
             decimals = 3) |>
  tab_header(title = "Play State Transition Probability Matrix",
             subtitle = "2016 Season")
```

<details>
<summary>table code</summary>
```{r, echo = TRUE, eval = FALSE}
P_matrix |>
  as_tibble(rownames = "state") |>
  filter(state == "000 0") |>
  pivot_longer(
    cols = -state, 
    names_to = "new_state", 
    values_to = "Prob" 
  ) |>
  filter(Prob > 0) |>
  arrange(desc(Prob)) |>
  gt() |>
  cols_align(align = "center") |>
  data_color(columns = Prob,
             palette = "inferno") |>
  fmt_number(columns = Prob,
             decimals = 3) |>
  tab_header(title = "Play State Transition Probability Matrix",
             subtitle = "2016 Season")
```
</details>

> starting from the “010 2” state, runner on second with two outs

```{r, echo = FALSE, eval = TRUE}
P_matrix |>
  as_tibble(rownames = "state") |>
  filter(state == "010 2") |>
  pivot_longer(
    cols = -state, 
    names_to = "new_state", 
    values_to = "Prob" 
  ) |>
  filter(Prob > 0) |>
  arrange(desc(Prob)) |>
  gt() |>
  cols_align(align = "center") |>
  data_color(columns = Prob,
             palette = "inferno") |>
  fmt_number(columns = Prob,
             decimals = 3) |>
  tab_header(title = "Play State Transition Probability Matrix",
             subtitle = "2016 Season")
```

<details>
<summary>table code</summary>
```{r, echo = TRUE, eval = FALSE}
P_matrix |>
  as_tibble(rownames = "state") |>
  filter(state == "010 2") |>
  pivot_longer(
    cols = -state, 
    names_to = "new_state", 
    values_to = "Prob" 
  ) |>
  filter(Prob > 0) |>
  arrange(desc(Prob)) |>
  gt() |>
  cols_align(align = "center") |>
  data_color(columns = Prob,
             palette = "inferno") |>
  fmt_number(columns = Prob,
             decimals = 3) |>
  tab_header(title = "Play State Transition Probability Matrix",
             subtitle = "2016 Season")
```
</details>


## Tracking Runs Scored

$$\text{runs} = (N_{\text{runners}}^{(b)} + O^{(b)} + 1) - (N_{\text{runners}}^{(a)} + O^{(a)})$$

* $N_{\text{runners}}$: number of runners in a state
* $O$: number of outs

```{r}
num_havent_scored <- function(s) {
  # INPUT: game state
  # OUTPUTS: number of runners and outs
  s |> str_split("") |> pluck(1) |> as.numeric() |> sum(na.rm = TRUE)}
```

```{r}
# apply to all possible states
runners_out <- T_matrix |> row.names() |> set_names() |> map_int(num_havent_scored)
```

```{r}
# for all possible pairs of states
R_runs <- outer(runners_out + 1, runners_out, 
  FUN = "-") |>           #difference in runs
  cbind("3" = rep(0, 24)) #ensure square matrix
```

```{r, echo = FALSE, eval = TRUE}
R_df <- R_runs |> 
  as.data.frame() 
R_df <- cbind(T_df$state, R_df)
colnames(R_df)[1] <- "state"

R_df |>
  gt() |>
  cols_align(align = "center") |>
  # data_color(columns = -state,
  #            palette = "inferno") |>
  tab_header(title = "Tracking Runs Scored",
             subtitle = "2016 Season")
```

<details>
<summary>table code</summary>
```{r, echo = TRUE, eval = FALSE}
R_df <- R_runs |> 
  as.data.frame() 
R_df <- cbind(T_df$state, R_df)
colnames(R_df)[1] <- "state"

R_df |>
  gt() |>
  cols_align(align = "center") |>
  # data_color(columns = -state,
  #            palette = "inferno") |>
  tab_header(title = "Tracking Runs Scored",
             subtitle = "2016 Season")
```
</details>


## Simulate Half-Inning

```{r}
simulate_half_inning <- function(P, R, start = 1) {
  # INPUTS:
      ## P: probability transition matrix
      ## R: run matrix
      ## start: starting state, 1 <= s <= 24
  # OUTPUT: number of runs scored
  s <- start
  path <- NULL
  runs <- 0
  while (s < 25) {
    s_new <- sample(1:25, size = 1, prob = P[s, ])
    path <- c(path, s_new)
    runs <- runs + R[s, s_new]
    s <- s_new
  }
  runs
}
```

### Many Iterations

```{r}
set.seed(20250611)
simulated_runs <- 1:12345 |>
  map_int(~simulate_half_inning(T_matrix, R_runs))
table(simulated_runs)
```

> How many runs are scored (on average) during a half-inning?

```{r}
mean(simulated_runs)
```

### All baserunner-outs states

```{r}
runs_j <- function(j) {
  1:987 |> #smaller number than textbook
    map_int(~simulate_half_inning(T_matrix, R_runs, j)) |>
    mean()
}

erm_2016_mc <- tibble(
  state = row.names(T_matrix), 
  mean_run_value = map_dbl(1:24, runs_j)
) |>
  mutate(
    bases = str_sub(state, 1, 3),
    outs_ct = as.numeric(str_sub(state, 5, 5))
  ) |>
  select(-state)
```


```{r, echo = FALSE, eval = TRUE}
erm_2016_mc |>
  pivot_wider(names_from = outs_ct, values_from = mean_run_value) |>
  gt() |>
  cols_align(align = "center") |>
  data_color(columns = -bases,
             palette = "inferno")  |>
  fmt_number(columns = -bases,
             decimals = 2) |>
  tab_header(title = "Runs Potential",
             subtitle = "2016 Season")
```

<details>
<summary>table code</summary>
```{r, echo = TRUE, eval = FALSE}
erm_2016_mc |>
  pivot_wider(names_from = outs_ct, values_from = mean_run_value) |>
  gt() |>
  cols_align(align = "center") |>
  data_color(columns = -bases,
             palette = "inferno")  |>
  fmt_number(columns = -bases,
             decimals = 2) |>
  tab_header(title = "Runs Potential",
             subtitle = "2016 Season")
```
</details>

> To understand the effect of non-batting plays (stealing, caught stealing, wild pitches, etc.) on run scoring, we compare this run expectancy matrix with the one found in Chapter 5 using all batting and non-batting plays. 

From Chapter 5, we previously computed an expected run matrix as `erm_2016`.

<details>
<summary>erm code</summary>
```{r, message = FALSE, warning = FALSE}
retro2016 <- retro2016 |>
  mutate(
    is_runner1 = as.numeric(
      run1_dest_id == 1 | bat_dest_id == 1
    ),
    is_runner2 = as.numeric(
      run1_dest_id == 2 | run2_dest_id == 2 | 
        bat_dest_id == 2
    ),
    is_runner3 = as.numeric(
      run1_dest_id == 3 | run2_dest_id == 3 |
        run3_dest_id == 3 | bat_dest_id == 3
    ),
    new_outs = outs_ct + event_outs_ct,
    new_bases = paste0(is_runner1, is_runner2, is_runner3),
    new_state = paste(new_bases, new_outs)
  )

retro2016 <- retro2016 |> 
  mutate(
    runs_before = away_score_ct + home_score_ct,
    half_inning = paste(game_id, inn_ct, bat_home_id),
    runs_scored = 
      (bat_dest_id > 3) + (run1_dest_id > 3) + 
      (run2_dest_id > 3) + (run3_dest_id > 3)
  )

half_innings <- retro2016 |>
  group_by(half_inning) |>
  summarize(
    outs_inning = sum(event_outs_ct), 
    runs_inning = sum(runs_scored),
    runs_start = first(runs_before),
    max_runs = runs_inning + runs_start
  )

retro2016 <- retro2016 |>
  inner_join(half_innings, by = "half_inning") |>
  mutate(runs_roi = max_runs - runs_before)

changes2016 <- retro2016 |> 
  filter(state != new_state | runs_scored > 0)

changes2016_complete <- changes2016 |>
 filter(outs_inning == 3)

erm_2016 <- changes2016_complete |> 
  group_by(bases, outs_ct) |>
  summarize(mean_run_value = mean(runs_scored)) |>
  ungroup()
```
</details>

```{r}
erm_2016_df <- erm_2016 |>
  inner_join(erm_2016_mc, join_by(bases, outs_ct)) |>
  mutate(
    run_value_diff = round(mean_run_value.x - mean_run_value.y, 2)
  ) |>
  select(bases, outs_ct, run_value_diff) |>
  pivot_wider(names_from = outs_ct, values_from = run_value_diff)
```

```{r, echo = FALSE, eval = TRUE}
erm_2016_df  |>
  gt() |>
  cols_align(align = "center") |>
  data_color(columns = -bases,
             palette = "inferno")  |>
  fmt_number(columns = -bases,
             decimals = 2) |>
  tab_header(title = "Value of Non-Batting Plays",
             subtitle = "2016 Season")
```

<details>
<summary>table code</summary>
```{r, echo = TRUE, eval = FALSE}
erm_2016_df  |>
  gt() |>
  cols_align(align = "center") |>
  data_color(columns = -bases,
             palette = "inferno")  |>
  fmt_number(columns = -bases,
             decimals = 2) |>
  tab_header(title = "Value of Non-Batting Plays",
             subtitle = "2016 Season")
```
</details>


## Stochastic Processes

### Multiple Transitions

> after three plate appearances

```{r}
P_matrix_3 <- P_matrix %*% P_matrix %*% P_matrix
```

```{r, echo = FALSE, eval = TRUE}
P_matrix_3 |>
  as_tibble(rownames = "state") |>
  filter(state == "000 0") |>
  pivot_longer(
    cols = -state, 
    names_to = "new_state", 
    values_to = "Prob" 
  ) |>
  filter(Prob > 0) |>
  arrange(desc(Prob)) |>
  gt() |>
  cols_align(align = "center") |>
  data_color(columns = Prob,
             palette = "inferno") |>
  fmt_number(columns = Prob,
             decimals = 3) |>
  tab_header(title = "After 3 Plate Appearances",
             subtitle = "starting with bases empty and zero outs")
```

<details>
<summary>table code</summary>

```{r, echo = TRUE, eval = FALSE}
P_matrix_3 |>
  as_tibble(rownames = "state") |>
  filter(state == "000 0") |>
  pivot_longer(
    cols = -state, 
    names_to = "new_state", 
    values_to = "Prob" 
  ) |>
  filter(Prob > 0) |>
  arrange(desc(Prob)) |>
  gt() |>
  cols_align(align = "center") |>
  data_color(columns = Prob,
             palette = "inferno") |>
  fmt_number(columns = Prob,
             decimals = 3) |>
  tab_header(title = "After 3 Plate Appearances",
             subtitle = "starting with bases empty and zero outs")
```
</details>

### Fundamental Matrix

$$N = (I - Q)^{-1}$$
```{r}
Q <- P_matrix[-25, -25] #without 3-out states
N <- solve(diag(rep(1, 24)) - Q) 
```

```{r}
N_0000 <- round(N["000 0", ], 2)
head(N_0000, n = 6)
```

Starting at the beginning of the inning (the “000 0” state)

* the average number of times the inning will be in the “000 0” state is 1.05
* the average number of times in the “000 1” state is 0.75
* the average number of times in the “000 2” state is 0.6
* etc.

```{r}
sum(N_0000)
```

* **the average number of plate appearances in a half-inning (before three outs)** is 4.27.


### Visit Frequency

> the length of the remainder of the inning, on average, starting with each possible state

$$N\vec{1}$$

```{r}
avg_num_plays <- N %*% rep(1, 24) |> t() |> round(2)
avg_num_plays[,1:8]
```

## For Individual Teams

```{r}
# retrieve team that is on offense (at bat)
retro2016_complete <- retro2016_complete |>
  mutate(home_team_id = str_sub(game_id, 1, 3),
         batting_team = if_else(bat_home_id == 0, away_team_id, home_team_id))

# tally states by team
T_team <- retro2016_complete |> group_by(batting_team, state, new_state) |>  count()
```

### Toward NOBLETIGER

```{r}
NOBL <- retro2016_complete |> 
  filter(state == "111 0") |>
  group_by(batting_team, state, new_state) |> tally() |> ungroup()
```

```{r, echo = FALSE, eval = TRUE}
NOBL |>
  group_by(batting_team) |>
  mutate(NOBL_sum = sum(n)) |>
  ungroup() |>
  select(batting_team, NOBL_sum) |>
  distinct() |>
  gt() |>
  cols_align(align = "center") |>
  data_color(columns = NOBL_sum,
             palette = "inferno") |>
  tab_header(title = "Toward NOBLETIGERs",
             subtitle = "count: no outs, bases loaded")
```

<details>
<summary>table code</summary>

```{r, echo = TRUE, eval = FALSE}
NOBL |>
  group_by(batting_team) |>
  mutate(NOBL_sum = sum(n)) |>
  ungroup() |>
  select(batting_team, NOBL_sum) |>
  distinct() |>
  gt() |>
  cols_align(align = "center") |>
  data_color(columns = NOBL_sum,
             palette = "inferno") |>
  tab_header(title = "Toward NOBLETIGERs",
             subtitle = "count: no outs, bases loaded")
```

</details>

> For some of the less common transitions, there is much variability in the counts across teams

### Smoothing Operation

$$p^{\text{EST}} = \frac{n}{n+K}p^{\text{TEAM}} + \frac{K}{n+K}p^{\text{ALL}}$$

* $n$: number of state transitions for TEAM
* $K$: smoothing parameter (e.g. $K = 1274$)

    - dependent of starting state
    

## Team Talent

### Bill James $\log_{5}$ model

* $P_{A}$: team A winning percentage
* $P_{B}$: team B winning percentage

$$P(\text{Team A wins}) = \frac{ \frac{P_{A}}{1-P_{A}}  }{ \frac{P_{A}}{1-P_{A}} + \frac{P_{B}}{1-P_{B}}}$$

* weighted average of *log odds* of winning

### Bradley-Terry Model

* $T_{A}$: team A talent level
* $T_{B}$: team B talent level

$$P(\text{Team A wins}) = \frac{ \exp(T_{A})  }{ \exp(T_{A})  + \exp(T_{B}) }$$

* similar to *softmax* formulas
* later: positive correlation

$$\text{more talent} \quad\Rightarrow\quad \text{more wins}$$

## Make a Schedule

```{r}
make_schedule <- function(teams, k) {
  # INPUTS:
      ## teams: vector of team names
      ## k: number of games played between two teams in the first team’s home park
  # OUTPUT: tibble with baseball season schedule
  num_teams <- length(teams)
  Home <- rep(rep(teams, each = num_teams), k)
  Visitor <- rep(rep(teams, num_teams), k)
  tibble(Home = Home, Visitor = Visitor) |>
    filter(Home != Visitor)
}
```

> to construct the schedule for the 1968 season

```{r}
teams_68 <- Lahman::Teams |>
  filter(yearID == 1968) |>
  select(teamID, lgID) |>
  mutate(teamID = as.character(teamID)) |>
  group_by(lgID)

schedule <- teams_68 |>
  group_split() |>
  set_names(pull(group_keys(teams_68), "lgID")) |>
  map(~make_schedule(teams = .x$teamID, k = 9)) |>
  list_rbind(names_to = "lgID")
```
"Note that schedule has $\frac{162 \cdot 20}{2}$ rows [permutations], since each game involves two teams.".  For example, we simulate the 1968 Detroit Tigers' schedule as

```{r, echo = FALSE, eval = TRUE}
schedule |>
  filter(Home == "DET" | Visitor == "DET") |>
  mutate(game_num = 1:162, .before = lgID) |>
  gt() |>
  cols_align(align = "center") |>
  tab_header(title = "Detroit Tigers 1968 Schedule",
             subtitle = "via simple construction") |>
  tab_style(
    locations = cells_body(columns = Home, rows = Home == "DET" ),
    style = list(cell_fill(color = "#FA4616"), cell_text(color = "#0C2340"))) |>
  tab_style(
    locations = cells_body(columns = Visitor, rows = Visitor == "DET" ),
    style = list(cell_fill(color = "#FA4616"), cell_text(color = "#0C2340"))) |>
  tab_style(
    locations = cells_body(columns = Home, rows = Home != "DET" ),
    style = list(cell_fill(color = "#0C2340"), cell_text(color = "#FA4616"))) |>
  tab_style(
    locations = cells_body(columns = Visitor, rows = Visitor != "DET" ),
    style = list(cell_fill(color = "#0C2340"), cell_text(color = "#FA4616")))
```

<details>
<summary>table code</summary>
```{r, echo = TRUE, eval = FALSE}
schedule |>
  filter(Home == "DET" | Visitor == "DET") |>
  mutate(game_num = 1:162, .before = lgID) |>
  gt() |>
  cols_align(align = "center") |>
  tab_header(title = "Detroit Tigers 1968 Schedule",
             subtitle = "via simple construction") |>
  tab_style(
    locations = cells_body(columns = Home, rows = Home == "DET" ),
    style = list(cell_fill(color = "#FA4616"), cell_text(color = "#0C2340"))) |>
  tab_style(
    locations = cells_body(columns = Visitor, rows = Visitor == "DET" ),
    style = list(cell_fill(color = "#FA4616"), cell_text(color = "#0C2340"))) |>
  tab_style(
    locations = cells_body(columns = Home, rows = Home != "DET" ),
    style = list(cell_fill(color = "#0C2340"), cell_text(color = "#FA4616"))) |>
  tab_style(
    locations = cells_body(columns = Visitor, rows = Visitor != "DET" ),
    style = list(cell_fill(color = "#0C2340"), cell_text(color = "#FA4616")))
```
</details>


## Compute Win Probabilities

Assume team talent is normally distributed

$$T \sim N(0, 0.2^{2})$$

```{r}
set.seed(20250611)
s_talent <- 0.20
teams_68 <- teams_68 |>
  mutate(talent = rnorm(10, 0, s_talent))

# increase chances of Cardinals' and Tigers' success
teams_68$talent[teams_68$teamID == "DET"] <- 0.35
teams_68$talent[teams_68$teamID == "SLN"] <- 0.31

schedule_talent <- schedule |>
  inner_join(teams_68, join_by(lgID, Home == teamID)) |>
  rename(talent_home = talent) |>
  inner_join(teams_68, join_by(lgID, Visitor == teamID)) |>
  rename(talent_visitor = talent)

schedule_talent <- schedule_talent |> 
  mutate(
    prob_home = exp(talent_home) /
      (exp(talent_home) + exp(talent_visitor))
  )
```

```{r, echo = FALSE, eval = TRUE}
set.seed(20250611)
schedule_talent |>
  filter(Home == "DET" | Visitor == "DET") |>
  slice_sample(n = 10) |>
  gt() |>
  cols_align(align = "center") |>
  fmt_number(columns = c(talent_home, talent_visitor, prob_home),
             decimals = 2) |>
  tab_header(title = "Detroit Tigers 1968 Schedule",
             subtitle = "sample of 10 games") |>
  tab_style(
    locations = cells_body(columns = c(Home, prob_home), rows = Home == "DET" ),
    style = list(cell_fill(color = "#FA4616"), cell_text(color = "#0C2340"))) |>
  tab_style(
    locations = cells_body(columns = Visitor, rows = Visitor == "DET" ),
    style = list(cell_fill(color = "#FA4616"), cell_text(color = "#0C2340"))) |>
  tab_style(
    locations = cells_body(columns = c(Home, prob_home), rows = Home != "DET" ),
    style = list(cell_fill(color = "#0C2340"), cell_text(color = "#FA4616"))) |>
  tab_style(
    locations = cells_body(columns = Visitor, rows = Visitor != "DET" ),
    style = list(cell_fill(color = "#0C2340"), cell_text(color = "#FA4616")))
```

<details>
<summary>table code</summary>
```{r, echo = TRUE, eval = FALSE}
set.seed(20250611)
schedule_talent |>
  filter(Home == "DET" | Visitor == "DET") |>
  slice_sample(n = 10) |>
  gt() |>
  cols_align(align = "center") |>
  fmt_number(columns = c(talent_home, talent_visitor, prob_home),
             decimals = 2) |>
  tab_header(title = "Detroit Tigers 1968 Schedule",
             subtitle = "sample of 10 games") |>
  tab_style(
    locations = cells_body(columns = c(Home, prob_home), rows = Home == "DET" ),
    style = list(cell_fill(color = "#FA4616"), cell_text(color = "#0C2340"))) |>
  tab_style(
    locations = cells_body(columns = Visitor, rows = Visitor == "DET" ),
    style = list(cell_fill(color = "#FA4616"), cell_text(color = "#0C2340"))) |>
  tab_style(
    locations = cells_body(columns = c(Home, prob_home), rows = Home != "DET" ),
    style = list(cell_fill(color = "#0C2340"), cell_text(color = "#FA4616"))) |>
  tab_style(
    locations = cells_body(columns = Visitor, rows = Visitor != "DET" ),
    style = list(cell_fill(color = "#0C2340"), cell_text(color = "#FA4616")))
```
</details>


## Simulate Season

```{r}
set.seed(20250611)
schedule_talent <- schedule_talent |>
  mutate(
    outcome = rbinom(nrow(schedule_talent), 1, prob_home),
    winner = if_else(outcome == 1, Home, Visitor)
)
```

```{r, echo = FALSE, eval = TRUE}
set.seed(20250611)
schedule_talent |>
  filter(Home == "DET" | Visitor == "DET") |>
  slice_sample(n = 10) |>
  gt() |>
  cols_align(align = "center") |>
  fmt_number(columns = c(talent_home, talent_visitor, prob_home),
             decimals = 2) |>
  tab_header(title = "Detroit Tigers 1968 Schedule",
             subtitle = "sample of 10 games") |>
  tab_style(
    locations = cells_body(columns = c(Home, prob_home), rows = Home == "DET" ),
    style = list(cell_fill(color = "#FA4616"), cell_text(color = "#0C2340"))) |>
  tab_style(
    locations = cells_body(columns = Visitor, rows = Visitor == "DET" ),
    style = list(cell_fill(color = "#FA4616"), cell_text(color = "#0C2340"))) |>
  tab_style(
    locations = cells_body(columns = c(Home, prob_home), rows = Home != "DET" ),
    style = list(cell_fill(color = "#0C2340"), cell_text(color = "#FA4616"))) |>
  tab_style(
    locations = cells_body(columns = Visitor, rows = Visitor != "DET" ),
    style = list(cell_fill(color = "#0C2340"), cell_text(color = "#FA4616")))
```

<details>
<summary>table code</summary>
```{r, echo = TRUE, eval = FALSE}
set.seed(20250611)
schedule_talent |>
  filter(Home == "DET" | Visitor == "DET") |>
  slice_sample(n = 10) |>
  gt() |>
  cols_align(align = "center") |>
  fmt_number(columns = c(talent_home, talent_visitor, prob_home),
             decimals = 2) |>
  tab_header(title = "Detroit Tigers 1968 Schedule",
             subtitle = "sample of 10 games") |>
  tab_style(
    locations = cells_body(columns = c(Home, prob_home), rows = Home == "DET" ),
    style = list(cell_fill(color = "#FA4616"), cell_text(color = "#0C2340"))) |>
  tab_style(
    locations = cells_body(columns = Visitor, rows = Visitor == "DET" ),
    style = list(cell_fill(color = "#FA4616"), cell_text(color = "#0C2340"))) |>
  tab_style(
    locations = cells_body(columns = c(Home, prob_home), rows = Home != "DET" ),
    style = list(cell_fill(color = "#0C2340"), cell_text(color = "#FA4616"))) |>
  tab_style(
    locations = cells_body(columns = Visitor, rows = Visitor != "DET" ),
    style = list(cell_fill(color = "#0C2340"), cell_text(color = "#FA4616")))
```
</details>

```{r}
results <- schedule_talent |> 
  group_by(winner) |>
  summarize(Wins = n()) |>
  inner_join(teams_68, by = c("winner" = "teamID"))
```


### Standings

```{r}
win_league <- function(res) {
  set.seed(20250611)
  res |>
    group_by(lgID) |>
    mutate(
      tiebreaker = runif(n = length(talent)),
      wins_total = Wins + tiebreaker,
      rank = min_rank(desc(wins_total)),
      is_winner_lg = wins_total == max(wins_total)
    )
}
sim_1968 <- win_league(results) |>
  rename(team = winner)
```

```{r, echo = FALSE, eval = TRUE}
sim_1968 |>
  arrange(rank) |>
  gt() |>
  cols_align(align = "center") |>
  data_color(columns = Wins,
             palette = "inferno") |>
  fmt_number(columns = c(talent, tiebreaker, wins_total),
             decimals = 2) |>
  tab_header(title = "1968 Season Standings",
             subtitle = "via simulation")
```

<details>
<summary>table code</summary>
```{r, echo = TRUE, eval = FALSE}
sim_1968 |>
  arrange(rank) |>
  gt() |>
  cols_align(align = "center") |>
  data_color(columns = Wins,
             palette = "inferno") |>
  fmt_number(columns = c(talent, tiebreaker, wins_total),
             decimals = 2) |>
  tab_header(title = "1968 Season Standings",
             subtitle = "via simulation")
```
</details>

### Simulate World Series

```{r}
set.seed(20250611)
ws_winner <- sim_1968 |>
  filter(is_winner_lg) |>
  ungroup() |>
  mutate(
    outcome = as.numeric(rmultinom(1, 7, exp(talent))),
    is_winner_ws = outcome > 3
  )
```

```{r, echo = FALSE, eval = TRUE}
ws_winner |>
  select(team, Wins, lgID, talent, outcome, is_winner_ws) |>
  gt() |>
  cols_align(align = "center") |>
  fmt_number(columns = talent,
             decimals = 2) |>
  tab_header(title = "1968 Post-Season",
             subtitle = "via simulation") |>
  tab_style(
    locations = cells_body(columns = c(team, outcome, is_winner_ws), 
                           rows = team == "DET" ),
    style = list(cell_fill(color = "#FA4616"), cell_text(color = "#0C2340"))) |>
  tab_style(
    locations = cells_body(columns = c(team, outcome, is_winner_ws), 
                           rows = team == "SLN" ),
    style = list(cell_fill(color = "#C41E3A"), cell_text(color = "#0C2340"))) |>
  tab_style(locations = cells_body(columns = outcome),
            style = cell_text(weight = "bold"))
```

<details>
<summary>table code</summary>
```{r, echo = TRUE, eval = FALSE}
ws_winner |>
  select(team, Wins, lgID, talent, outcome, is_winner_ws) |>
  gt() |>
  cols_align(align = "center") |>
  fmt_number(columns = talent,
             decimals = 2) |>
  tab_header(title = "1968 Post-Season",
             subtitle = "via simulation") |>
  tab_style(
    locations = cells_body(columns = c(team, outcome, is_winner_ws), 
                           rows = team == "DET" ),
    style = list(cell_fill(color = "#FA4616"), cell_text(color = "#0C2340"))) |>
  tab_style(
    locations = cells_body(columns = c(team, outcome, is_winner_ws), 
                           rows = team == "SLN" ),
    style = list(cell_fill(color = "#C41E3A"), cell_text(color = "#0C2340"))) |>
  tab_style(locations = cells_body(columns = outcome),
            style = cell_text(weight = "bold"))
```
</details>


## Simulate Many Seasons

We can wrap the previous codes into a `simulate_1968` function

```{r, echo = TRUE, eval = FALSE}
simulate_1968 <- function(s_talent = 0.20){ ... }
```

<details>
<summary>function code</summary>
```{r}
simulate_1968 <- function(s_talent = 0.20){
  teams_68 <- teams_68 |>
    mutate(talent = rnorm(10, 0, s_talent))
  
  # increase chances of Cardinals' and Tigers' success
  teams_68$talent[teams_68$teamID == "DET"] <- rnorm(1, 0.3, 0.05)
  teams_68$talent[teams_68$teamID == "SLN"] <- rnorm(1, 0.27, 0.04)
  
  schedule_talent <- schedule |>
    inner_join(teams_68, join_by(lgID, Home == teamID)) |>
    rename(talent_home = talent) |>
    inner_join(teams_68, join_by(lgID, Visitor == teamID)) |>
    rename(talent_visitor = talent)
  
  schedule_talent <- schedule_talent |> 
    mutate(
      prob_home = exp(talent_home) /
        (exp(talent_home) + exp(talent_visitor))
    )
  
  schedule_talent <- schedule_talent |>
    mutate(
      outcome = rbinom(nrow(schedule_talent), 1, prob_home),
      winner = if_else(outcome == 1, Home, Visitor)
    )
  
  results <- schedule_talent |> 
    group_by(winner) |>
    summarize(Wins = n()) |>
    inner_join(teams_68, by = c("winner" = "teamID"))
}
```
</details>

and use the `rep` function to simulate 100 copies of the 1968 season.

```{r}
set.seed(20250611)
many_results <- rep(0.20, 100) |>
  map(simulate_1968) |>
  list_rbind() |>
  rename(team = winner)
```

```{r, echo = FALSE, eval = TRUE}
title_text <- "<span style='color:#0C2340'>Simulations of **1968 MLB Season**</span>"
subtitle_text <- "with focus on the highly talented <span style='color:#FA4616'>**Detroit Tigers**</span> team"

many_results |>
  ggplot() +
  geom_point(aes(x = talent, y = Wins),
             alpha = 0.2,
             color = "#0C2340",
             data = many_results |>
               filter(team != "DET")) +
  geom_point(aes(x = talent, y = Wins),
             alpha = 0.6,
             size = 3,
             color = "#FA4616",
             data = many_results |>
               filter(team == "DET")) +
  labs(title = title_text,
       subtitle = subtitle_text,
       caption = "DSLC") +
  theme_minimal() +
  theme(plot.subtitle = ggtext::element_markdown(),
        plot.title = ggtext::element_markdown()) +
  xlim(-1,1)
```

<details>
<summary>graph code</summary>
```{r, echo = TRUE, eval = FALSE}
title_text <- "<span style='color:#0C2340'>Simulations of **1968 MLB Season**</span>"
subtitle_text <- "with focus on the highly talented <span style='color:#FA4616'>**Detroit Tigers**</span> team"

many_results |>
  ggplot() +
  geom_point(aes(x = talent, y = Wins),
             alpha = 0.2,
             color = "#0C2340",
             data = many_results |>
               filter(team != "DET")) +
  geom_point(aes(x = talent, y = Wins),
             alpha = 0.6,
             size = 3,
             color = "#FA4616",
             data = many_results |>
               filter(team == "DET")) +
  labs(title = title_text,
       subtitle = subtitle_text,
       caption = "DSLC") +
  theme_minimal() +
  theme(plot.subtitle = ggtext::element_markdown(),
        plot.title = ggtext::element_markdown()) +
  xlim(-1,1)
```
</details>

* average team $T = 0$ tend to win about 81 games.
* positive correlation: more talent $\Rightarrow$ more wins


### Parity

> What is the relationship between a team’s talent and its post-season success?

```{r}
set.seed(20250611)
many_results <- rep(0.20, 100) |>
  map(abdwr3edata::one_simulation_68) |>
  list_rbind()
```

$$P(\text{win championship}|T) = \frac{e^{a+bT}}{1 + e^{a+bT}}$$

```{r}
fit1 <- glm(
  Winner.Lg ~ Talent,
  data = many_results, family = binomial
)
fit2 <- glm(
  Winner.WS ~ Talent,
  data = many_results, family = binomial
)
```

```{r, echo = FALSE, eval = TRUE}
tdf <- tibble(
  Talent =  seq(-0.4, 0.4, length.out = 100)
)

title_text <- "**Championship Probability** versus **Team Talent**"
subtitle_text <- "<span style='color:#0C2340'>**Pennant Probabilities**</span> and <span style='color:#FA4616'>**World Series Probabilities**</span>"

tdf |>
  mutate(
    Pennant = predict(fit1, newdata = tdf, type = "response"),
    `World Series` = predict(fit2, newdata = tdf, type = "response")
  ) |>
  pivot_longer(
    cols = -Talent,
    names_to = "Outcome", 
    values_to = "Probability"
  ) |>
  ggplot(aes(Talent, Probability, color = Outcome)) +
  geom_line(linewidth = 3) +
  labs(title = title_text,
       subtitle = subtitle_text,
       caption = "DSLC") +
  scale_color_manual(values = c("#0C2340", "#FA4616")) +
  theme_minimal() +
  theme(legend.position = "none",
        plot.subtitle = ggtext::element_markdown(hjust = 1.0),
        plot.title = ggtext::element_markdown(hjust = 1.0)) +
  ylim(0, 0.65)
```

<details>
<summary>graph code</summary>
```{r, echo = TRUE, eval = FALSE}
tdf <- tibble(
  Talent =  seq(-0.4, 0.4, length.out = 100)
)

title_text <- "**Championship Probability** versus **Team Talent**"
subtitle_text <- "<span style='color:#0C2340'>**Pennant Probabilities**</span> and <span style='color:#FA4616'>**World Series Probabilities**</span>"

tdf |>
  mutate(
    Pennant = predict(fit1, newdata = tdf, type = "response"),
    `World Series` = predict(fit2, newdata = tdf, type = "response")
  ) |>
  pivot_longer(
    cols = -Talent,
    names_to = "Outcome", 
    values_to = "Probability"
  ) |>
  ggplot(aes(Talent, Probability, color = Outcome)) +
  geom_line(linewidth = 3) +
  labs(title = title_text,
       subtitle = subtitle_text,
       caption = "DSLC") +
  scale_color_manual(values = c("#0C2340", "#FA4616")) +
  theme_minimal() +
  theme(legend.position = "none",
        plot.subtitle = ggtext::element_markdown(hjust = 1.0),
        plot.title = ggtext::element_markdown(hjust = 1.0)) +
  ylim(0, 0.65)
```
</details>


