# Home Run Hitting

```{r, include = FALSE }
library(abdwr3edata)
library(mgcv)
library(metR)
library(tidyverse)
library(arrow)
```

**Learning objectives:**

- Examine some of the factors that determine home runs:
   - Launch angle and Exit Velocity
   - Spray Angle
   - Park effects
   
- Compare 2021 and 2023 season

[Chapter reference](https://beanumber.github.io/abdwr3e/13-homeruns.html)

## Getting the Data {-}

- Book uses a data file called 'sc_bip_2021_2023.rds' which is not included. 

- Following the guidance of chapter 21, it can be created 
   - Pull down statcast data (takes some time!)
   - Filter to BIP and reduce columns
   - Combine and save as RDS file.

## Code for creating the data file {-}

```{r, eval = FALSE}
library(abdwr3edata)
library(tidyverse)
library(fs)
library(arrow)
data_dir <- here::here("data_large")
statcast_dir <- path(data_dir, "statcast_csv")
if (!dir.exists(statcast_dir)) {
  dir.create(statcast_dir)
}
statcast_season(year = 2023, dir = statcast_dir)
statcast_season(year = 2021, dir = statcast_dir)
sc2023 <- statcast_dir |> statcast_read_csv(pattern = "2023.+\\.csv")
sc2021 <- statcast_dir |> statcast_read_csv(pattern = "2021.+\\.csv")
sc2021_bip <- sc2021 |> filter(type == "X") 
sc2023_bip <- sc2023 |> filter(type == "X") 
sc_bip_2021_2023 <- bind_rows(sc2021_bip, sc2023_bip) |>
                    select(game_pk, game_date, batter, pitcher, events, 
                          stand, p_throws, hit_distance_sc, hc_x, hc_y, 
                          launch_speed, launch_angle, home_team, away_team, Season, HR)
write_parquet(sc_bip_2021_2023, "data/sc_bip_2021_2023.parquet")
```

## Read in data {-}

```{r}
sc_two_seasons <- here::here("data/sc_bip_2021_2023.parquet") |>
  read_parquet() |> 
  mutate(
    Season = year(game_date),
    HR = ifelse(events == "home_run", 1, 0)
  )
sc_2023 <- sc_two_seasons |> 
  filter(Season == 2023)

```

```
Rows: 124,234
Columns: 16
$ game_pk         <dbl> 718773, 718774, 718773, 718778, 718781, 718778, 718772, 718770, 718776, 718778, 718…
$ game_date       <date> 2023-03-30, 2023-03-30, 2023-03-30, 2023-03-30, 2023-03-30, 2023-03-30, 2023-03-30…
$ batter          <dbl> 613564, 643446, 641584, 453568, 527038, 592178, 665489, 502110, 670623, 602074, 663…
$ pitcher         <dbl> 656605, 645261, 656605, 605483, 543037, 605483, 571945, 668678, 593958, 605483, 669…
$ events          <chr> "triple", "single", "single", "single", "grounded_into_double_play", "field_out", "…
$ stand           <chr> "L", "L", "L", "L", "R", "R", "R", "R", "R", "R", "L", "L", "R", "L", "L", "R", "R"…
$ p_throws        <chr> "R", "R", "R", "L", "R", "L", "R", "R", "L", "L", "R", "R", "L", "R", "R", "R", "R"…
$ hit_distance_sc <dbl> 134, 9, 254, 162, 51, 56, 42, 185, 143, 171, 240, 374, 356, 65, 392, 116, 422, 188,…
$ hc_x            <dbl> 215.08, 164.78, 196.95, 90.66, 110.24, 153.95, 184.78, 177.78, 116.88, 177.33, 73.8…
$ hc_y            <dbl> 107.23, 105.10, 95.17, 133.88, 148.44, 209.01, 100.21, 106.58, 74.73, 114.14, 117.8…
$ launch_speed    <dbl> 94.2, 93.7, 111.7, 59.1, 94.8, 69.5, 115.5, 102.7, 105.7, 93.8, 93.4, 96.4, 93.5, 8…
$ launch_angle    <dbl> 9, -19, 13, 27, 1, 81, -2, 9, 7, 9, 55, 28, 25, 76, 35, 6, 26, 12, 9, 32, -2, 1, -3…
$ home_team       <chr> "CIN", "MIA", "CIN", "SD", "NYY", "SD", "STL", "LAD", "TB", "SD", "SEA", "MIA", "TB…
$ away_team       <chr> "PIT", "NYM", "PIT", "COL", "SF", "COL", "TOR", "AZ", "DET", "COL", "CLE", "NYM", "…
$ Season          <dbl> 2023, 2023, 2023, 2023, 2023, 2023, 2023, 2023, 2023, 2023, 2023, 2023, 2023, 2023,…
$ HR              <dbl> 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0,…
```


- Parquet format allows us to include the data in the repo! (3.4 MB)

## Home runs and launch variables {-}

$$
\log(\frac{\text{Pr}(HR)}{1-\text{Pr}(HR)}) = s(LA, LS)
$$
- Logistic GAM regression of probability of a Home Run vs:

   - LA : Launch angle
   
   - LS : Launch speed
   
- `s` is used in GAM to indicate a smooth function of the two measurements.

```{r}
library(mgcv)
fit_23 <- gam(
  HR ~ s(launch_angle, launch_speed),
  family = binomial,
  data = sc_2023
)
```


We can get predictions from the model:

```{r}
fit_23 |>
  predict(
    newdata = data.frame(
      launch_speed = 105, 
      launch_angle = 25
    ),
    type = "response"
  )
```

## Plot of model  {-}

```{r}
grid <- expand_grid(
  launch_angle = seq(15, 40, length = 50),
  launch_speed = seq(90, 110, length = 50)
)
hats <- fit_23 |>
  predict(newdata = grid, type = "response")
grid <- grid |>
  mutate(prob = hats)

ggplot(grid) +
  geom_contour_fill(
    aes( x = launch_angle, y = launch_speed, z = prob),
    breaks = c(0, .1, .3, .5, .7, .9, 1),
    linewidth = 1.5
  ) +
  scale_fill_viridis_c(option = "H") + 
  theme(text = element_text(size = 18)) +
  labs(x = "Launch Angle", y = "Launch Speed") +
  guides(fill = guide_legend(title = "Prob (HR)"))
```

## Optimal launch angle? {-}

- Previous graph suggests:
   - Higher launch speed always better
   - But for launch angle, there is an optimal
- Focus on 100-105 mph balls:

```{r}
sc_2023 |>
  filter(launch_speed >= 100, launch_speed <= 105) |>
  ggplot(aes(x = launch_angle, y = HR)) + geom_smooth(method = "gam") +
      scale_y_continuous( "Probability of a Home Run", limits = c(0, 1)) +
      scale_x_continuous( "Launch Angle (degrees)", limits = c(10, 50))
```
 
## Temperature effects {-}

- Conventional wisdom: warmer temps -> more home runs. 

- Add in temperature data (included in `abdwr3edata` package)

- Also need to focus on outside parks, so include park data (also in package)

```{r}
temps_parks_2023 <- temps_2023 |> inner_join(parks_2023, by = c("Park"))
sc_2023 <- sc_2023 |> inner_join(temps_parks_2023, by = "game_pk")
temp_hr <- filter(sc_2023, Dome == "No") |> 
  group_by(temperature) |> 
  summarize(
    BIP = n(),
    HR = sum(HR, na.rm = TRUE)
  ) |> 
  mutate(HR_Rate = 100 * HR / BIP)
```

```{r, echo=FALSE}
temp_hr |>
  filter(temperature >= 55, temperature <= 90) |>
  ggplot(aes(temperature, HR_Rate)) +
  geom_point() +
  geom_smooth(method = "lm", formula = "y ~ x") +
  labs(x = "Temperature (deg F)", y = "Home Run Rate")
```

- Seems to hold up. A linear fit yields about 0.04% increase per degree (4.65% at 70)

## Spray angle {-}

```{r}
sc_2023 <- sc_2023 |> 
  mutate(location_x = 2.5 * (hc_x - 125.42),
         location_y = 2.5 * (198.27 - hc_y),
         spray_angle = atan(location_x / location_y) * 180 / pi
  )
sc_hr <- sc_2023 |> filter(events == "home_run")

ggplot(sc_hr, aes(spray_angle, hit_distance_sc)) +
  geom_point(alpha = 0.25)
```

## Home runs spray vs batting side {-}

```{r}
ggplot(sc_hr, aes(spray_angle)) +
  geom_density() +
  facet_wrap(vars(stand), ncol = 1)
```

- Confirms that batters tend to hit home runs in "pull" direction.
- Also note that Left handers tend to pull stronger!

## Ball park effects (reprise) {-}

- Compute home and away HRs for each team
- Compute park factor as ratio of home HRs to away HRs.

```{r}
sc_home <- sc_2023 |> 
  group_by(home_team) |> 
  summarize(HR = sum(HR))
sc_away <- sc_2023 |> 
  group_by(away_team) |> 
  summarize(HR = sum(HR))
pf <- sc_home |>
  inner_join(sc_away, join_by(home_team == away_team)) |> 
  mutate(Park_Factor = HR.x / HR.y)
```

## Park Factor Plot {-}

```{r, echo= FALSE}
ggplot(pf, aes(Park_Factor, home_team)) +
  geom_point() +
  geom_vline(xintercept = 1, color = "blue")
```

- Blue line indicates 'neutral'
- Standouts: TEX and NYY > 1.5 ! 

## Pitcher or batter ? {-}

- Are home runs about the hitter or pitcher?

- One way to address: use random effects model. 

$$
\log \left(\frac{p}{1-p}\right) = \mu + \beta_i + \gamma_j
$$

-  $\mu$ is the overall effect
-  $\beta_i$ is the effect due to the $i$th hitter 
-  $\gamma_j$ is the effect due to the $j$th pitcher. 
-  $\beta$ and $\gamma$ are both assumed to be normally distributed:
    - $\beta_i \sim N(0,\sigma_h)$
    - $\gamma_j \sim N(0,\sigma_p)$

## Fitting the model {-}

- In `lme4` (and others) this is expressed as: 


```{r, warning=FALSE, message = FALSE}
library(lme4)
fit <- glmer(
  HR ~ (1 | pitcher) + (1 | batter),
  data = sc_2023,
  family = binomial
)
VarCorr(fit)  # get sigma_p and sigma_b
```

-  Results suggest more of the variation in home runs is due to batter effects then pitcher effects. 

## Comparing Home run hitting accross seasons {-}

- Home run hitting in 2021 and 2023 seem similar: about 1.2 home runs per game
- However, there might be differences when considering launch properties
- Approach:
    - Bin the launch variable space (LA / LS) 
    - Examine changes in home runs / balls in play / rates of both
    
## Definitions {-}

$$
BIP \, Rate = 100 \times \frac{BIP}{N},
$$

$$
HR \, Rate = 100 \times \frac{HR}{BIP}.
$$

## Binning launch variables {-}

```{r}
bin_rates <- function(sc_ip, LA_breaks, LS_breaks) {
  Total_BIP <- nrow(sc_ip)
  sc_ip |> 
    mutate(
      LS = cut(launch_speed, breaks = LS_breaks),
      LA = cut(launch_angle, breaks = LA_breaks)
    ) |> 
    filter(!is.na(LA), !is.na(LS)) |> 
    group_by(LA, LS) |> 
    summarize(
      BIP = n(),
      HR = sum(HR),
      .groups = "drop"
    ) |> 
    mutate(
      BIP_Rate = 100 * BIP / Total_BIP,
      HR_Rate = 100 * HR / BIP
    ) 
}
```


## Function to plot results {-}
```{r}
compute_bin_midpoint <- function(x) {
  x |>
    as.character() |>
    str_split_1(",") |>
    map_dbl(parse_number) |>
    mean()
}

bin_plot <- function(S, LA_breaks, LS_breaks, label) {
  S |> 
    mutate(
      la = map_dbl(LA, compute_bin_midpoint),
      ls = map_dbl(LS, compute_bin_midpoint)
    ) |>
    ggplot(aes(x = la, y = ls)) + 
    geom_text(aes(label = {{label}}), size = 8) +
    geom_vline(
      xintercept = LA_breaks,
      color = "blue"
    ) +
    geom_hline(
      yintercept = LS_breaks,
      color = "blue"
    ) +
    theme(text = element_text(size = 18)) +
    labs(x = "Launch Angle", y = "Launch Speed") 
}
```

## Example 2023 HR by bin {-}

```{r}
LA_breaks <- seq(20, 40, by = 5)
LS_breaks <- seq(95, 110, by = 5)

S <- sc_2023 |>
  bin_rates(LA_breaks, LS_breaks)

bin_plot(S, LA_breaks, LS_breaks, HR)
```

## 2023 HR Rate by Bin {-}

```{r}
bin_plot(S, LA_breaks, LS_breaks, label = paste(round(HR_Rate, 0), "%", sep = ""))
```

## Compare between seasons {-}

```{r}
S2 <- sc_two_seasons |>
  group_split(Season) |>
  map(bin_rates, LA_breaks, LS_breaks) |>
  set_names(c(2021, 2023)) |>
  list_rbind(names_to = "year")
```


- Compare rates on logit scale to avoid endpoint issues


```{r}
logit <- function(x){
  log(x) - log(100 - x)
}

d_BIP <- S2 |>
  select(year, LA, LS, BIP_Rate) |>
  pivot_wider(
    names_from = year, names_prefix = "y",
    values_from = BIP_Rate
  ) |>
  mutate(d_BIP_logits = logit(y2023) - logit(y2021))
```

## Difference in BIP logits {-}

```{r}
  d_BIP |>
  bin_plot(
    LA_breaks, LS_breaks, 
    label = round(d_BIP_logits, 2)
  )
```

- Are these significant? Question for reader ;)

## Changes in carry? {-}

- Given launch variables, how has probability BIP becomes HR changed?


```{r}
S2 |>
  select(year, LA, LS, HR_Rate) |>
  pivot_wider(
    names_from = year, names_prefix = "y",
    values_from = HR_Rate
  ) |>
  mutate(d_HR_logits = logit(y2023) - logit(y2021)) |>
  bin_plot(
    LA_breaks, LS_breaks, 
    label = round(d_HR_logits, 2)
  )
```

- Biggest change is in 30-35 launch angle.  

## Interpretation {-}

- Note :

$$
\begin{aligned}
d_{logit} &= \log \left( \frac{R_{2023}}{100 - R_{2023}}\right) - \log \left( \frac{R_{2021}}{100 - R_{2021}}\right) \\
          &= \log \left( \frac{R_{2023} / (100 - R_{2023})} {R_{2021}/(100 - R_{2021})}\right) \\
\end{aligned}
$$

- $d_{logit} = log(\text{odds_ratio})$

- So for example a decrease of $d_{logit}$ of -.61 corresponds to an 'odds' ratio:

```{r}
exp(-.61)
```
- So the odds went down by about a factor of 2.

- Book claims this is compensated by more balls hit hard.




